Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-24 main.tex
Edits in the methods section.

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  has a stabilizing effect down to about 1~\si{\kph}. We hypothesize that a
  stabilized bicycle will reduce the probability of falling. To test the
  motor's possible assistance during falls, we apply varying magnitude external
  handlebar perturbations to twenty-six subjects who rode on a treadmill with
  the balance assist system activated and deactivated. The probability of not
  recovering from a handlebar perturbation significantly decreases when the
  balance assist is activated at a travel speed of 6~\si{\kph}. This positive
  effect is most prominent at and around the individual riders' perturbation
  resistance threshold. We conclude that use of a balance assist system in real
  world bicycling would reduce the number of falls that occur near riders'
  control authority limits.
}

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. In the speed range tested, the commanded
steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
proportional feedback gain linearly with respect to speed stabilizes the
normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the bicycle without the inertial effect of a rigid rider (left) and
    with a rigid rider (right). Vertical dotted and dotted-dashed lines
    indicate the two speeds we perform experiments at. Grey shaded region is
    the linear stable speed range. Geometric and inertial parameters for these
    plots were estimated with the methods presented in \cite{Moore2012} and
    software packages BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon
    1.5.0~\citep{Dembia2015} and are shown in
    Appendix~\ref{sec:physical-parameters}.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from Earth fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four motors working in cooperation, Figure~\ref{fig:setup-diagram}. The four
motors are programmed to apply a light force at all times to keep the ropes
taught and to track a commanded force profile using a PID controller running on
a microprocessor (Arduino Mega 2560, Arduino, Italy). We measure the force
applied by each motor at the handlebar via four inline 250~\si{\newton} S-style
load cells (Bosche GmbH, Damme, Germany) except the left rear load cell that
has a max load of 500~\si{\newton}. The commanded force profiles are designed
to apply an external pulsive torque to the front assembly (handlebars, forks,
wheel) at magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four
motors (EC-90 Flat, Maxon Group, Switzerland) are arranged at the four corners
of a 1~\si{\meter} wide treadmill (Bonte Technology B.V., Zwolle, The
Netherlands) that can reach speeds of 18~\si{\kilo\meter\per\hour}. The general
design of this system is described in detail in Van De Velde's MSc
thesis~\citep{vandeVelde2022a} and the physical arrangement is shown in
Figure~\ref{fig:participant-in-set-up}. Our modifications relative to Van De
Velde include simplifying the Bump'Em motor controller with an inexpensive
microcontroller and the use of a simpler non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

\subsection{Protocol}
%
We recruited 26 able-bodied young adults (20-36 years old) to participate in
the experiments.
\todo[]{Include the average age, mass, height and standard deviation if we
recorded those.}
The subjects were all confident in their cycling skills and had cycled at least
once in the last month. Eleven subjects performed the experiments at
6~\si{\kilo\meter\per\hour} (1.7~\si{\meter\per\second}) and fifteen subjects
performed the experiments at 10~\si{\kilo\meter\per\hour}
(2.8~\si{\meter\per\second}). All subjects consented to the experiment and
could decline to continue at any time. The study was approved by Delft
University of Technology's Human Research Ethics Council (\#3897).

The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 10~\si{\kilo\meter\per\hour} with the controller
gain set to $g=8$ and the second group performed the protocol at a belt speed
of 6~\si{\kilo\meter\per\hour} with the controller gain set to $g=10$. The
10~\si{\kph} group's experiments occurred two weeks before the
6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, while attempting to
maintain the location of their front wheel on the center line of the treadmill.
We define a ``fall'' on the treadmill by two criteria: 1) the rider removes
their foot from the pedal and places it on the ground or 2) the bicycle wheel
exceeds the width of the treadmill belt. We then applied perturbations in
random directions (clockwise or counter-clockwise), starting at 20~\si{\newton}
and increasing the magnitude by 30~\si{\newton} until the participants
fell. Figure~\ref{fig:perturbation-sequence} shows an example resulting motion
from a perturbation. We log the magnitude that causes the first fall to
characterize that subject's \emph{nominal fall threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

We record the order number of each perturbation, i.e. first, second, third,
\ldots, to measure how many perturbations a rider is subjected to before the
current pertubration. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain setting on the balance assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for "fall". All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \cite{Enders2007}, with clusters being an
individual subject. The clusters are chosen as all data from an individual
subject because we are only interested in the association between the state of
the balance assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation. However, this was not true. Cluster-mean centering
shows there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
set of data. Statistics were computed with the lme4 version
1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.

@@

\subsection{Application of the Logistic Model}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain some
insight.
But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall:
\(e^{\alpha_k}=0.53\).
This means if the odds are a 1000:1, turning on the balance assist system
reduces the odds to 500:1.
However, that case the probability that a fall occurs is only reduced from
$0.999$ to $0.998$.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse the rider was subjected to.
This means that the balance assist is most effective for perturbations close to
the subject's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

@@

\section*{Acknowledgements}
%
This study follows and draws from experimental and analysis methods originally
developed in unpublished research by Marco Reijne. The authors acknowledge
Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
Reijne, and Arend Schwab for contributions to the development of the
balance assist bicycle. We thank all of the assistants that helped run the
experiments and the subjects for their time an enthusiasm. We thank Peter
Stahlecker for valuable feedback in early revisions.

>>>>

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add an
  automatically controlled steering motor to a consumer electric bicycle that
  has a stabilizing effect down to about 3~\si{\kph}. We hypothesize that a
  stabilized bicycle will reduce the probability of falling. To test the
  motor's possible assistance during falls, we apply varying magnitude external
  handlebar perturbations to twenty-six participants who rode on a treadmill with
  the balance assist system activated and deactivated. The probability of not
  recovering from a handlebar perturbation significantly decreases when the
  balance assist is activated at a travel speed of 6~\si{\kph}. This positive
  effect is most prominent at and around the individual riders' perturbation
  resistance threshold. We conclude that use of a balance assist system in real
  world bicycling would reduce the number of falls that occur near riders'
  control authority limits.
}

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) to tune the gain magnitude during the experiments.
Scaling the proportional feedback gain linearly with respect to speed
stabilizes the normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the bicycle without the inertial effect of a rigid rider (left) and
    with a rigid rider (right). Vertical dotted and dotted-dashed lines
    indicate the two speeds we perform experiments at. Grey shaded region is
    the linear stable speed range. Geometric and inertial parameters for these
    plots were estimated with the methods presented in \citet{Moore2012} and
    software packages BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon
    1.5.0~\citep{Dembia2015} and are shown in
    Appendix~\ref{sec:physical-parameters}.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from Earth fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four motors (EC-90 Flat, Maxon Group, Switzerland) working in cooperation,
Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
force at all times to keep the ropes taught and to track a commanded force
profile using a PID controller running on a microprocessor (Arduino Mega 2560,
Arduino, Italy). We measure the force applied by each motor at the handlebar
via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
rated for 250~\si{\newton} except the left rear load cell that was rated for
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front assembly (handlebars, forks, wheel) at
magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four
motors are arranged at the four corners of a 1~\si{\meter} wide treadmill
(Bonte Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
18~\si{\kilo\meter\per\hour}. The general design of this perturbation system is
described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde's design include simplifying the
Bump'Em motor controller with an inexpensive microcontroller and the use of a
simpler non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

\subsection{Protocol}
%
We recruited 26 able-bodied young adults (20-36 years old) to participate in
the experiments. The participants were all confident in their cycling skills and
had cycled at least once in the last month. All participants consented to the
experiment and could decline to continue at any time. The study was approved by
Delft University of Technology's Human Research Ethics Council (\#3897).

The participants were divided into two groups. The first group of fifteen
participants
performed the protocol at a belt speed of
10~\si{\kilo\meter\per\hour}(2.8~\si{\meter\per\second}) with the controller
gain set to $g=8$ and the second group of eleven particpants performed the
protocol at a belt speed of 6~\si{\kilo\meter\per\hour}
(1.7~\si{\meter\per\second}) with the controller gain set to $g=10$. The
10~\si{\kph} group's experiments occurred two weeks before the
6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and they were attached to the ceiling via a fall safety
harness, Figure~\ref{fig:participant-in-set-up}. We allowed them to practice
riding on the treadmill until they indicated they were comfortable enough to
have perturbations applied. For most, this was less than a 10~\si{\minute} warm
up.  We then asked the rider to ride for 90~\si{\second}, while attempting to
maintain the location of their front wheel on the center line of the treadmill
as a final preparation before the perturbations.  We then applied perturbations
in random directions (clockwise or counter-clockwise), starting at
20~\si{\newton} and increasing the magnitude by 30~\si{\newton} until the
participants fell. We define a ``fall'' on the treadmill by two criteria: 1)
the rider removes their foot from the pedal and places it on the ground or 2)
the bicycle wheel exceeds the width of the treadmill belt.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We log the magnitude that causes the first fall to characterize
that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

We record the order number of each perturbation, i.e. first, second, third,
\ldots, to measure how many perturbations a rider is subjected to before the
current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain setting on the balance assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for "fall". All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \cite{Enders2007}, with clusters being an
individual participant. The clusters are chosen as all data from an individual
participant because we are only interested in the association between the state of
the balance assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation. However, this was not true. Cluster-mean centering
showed there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
set of data. Statistics were computed with the lme4 version
1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.

@@

\subsection{Application of the Logistic Model}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain some
insight.
But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall:
\(e^{\alpha_k}=0.53\).
This means if the odds are a 1000:1, turning on the balance assist system
reduces the odds to 500:1.
However, that case the probability that a fall occurs is only reduced from
$0.999$ to $0.998$.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse the rider was subjected to.
This means that the balance assist is most effective for perturbations close to
the participant's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

@@

\section*{Acknowledgements}
%
This study follows and draws from experimental and analysis methods originally
developed in unpublished research by Marco Reijne. The authors acknowledge
Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
Reijne, and Arend Schwab for contributions to the development of the balance
assist bicycle. We acknowledge Shannon van de Velde's and Marco Reijne's
development of the bicycle perturbation system. We thank all of the assistants
that helped run the experiments and the participants for their time an enthusiasm.
We thank Peter Stahlecker for valuable feedback in early revisions.
